Hash Bash: Genetic Algorithms for Hash Function Generation

### 2008-12-03 paper/paper.tex
paper almost proof-read. need to put together the citations page.

\paragraph{}
The genetic algorithm is inspired by biological natural selection. John Holland of the University of Michigan pioneered its use in the 1970s (Holland, “Genetic algorithms and the optimal allocation of trials.” SIAM J. Comput., 2.88-105); however, it wasn't until the 1980s and 1990s that the computational capacity for practical implementations of the algorithm became readily available. The algorithm as we use it consists of a series of functions - a randomized generator, a fitness evaluator, and a breeder - that operate on chromomsomes, which are strings formed from a genetic description alphabet. The algorithm proceeds as following:
\begin{enumerate}
	\item Create a random starting population of size S.
	\item Evaluate the fitness of every member of the current generation.
	\item Breed the fittest members to form the next generation.
	\item Add random or old members until we have the next generation is once more of size S.
	\item Go to step 2 if we're not done.
\end{enumerate}

@@

\begin{itemize}
\item No registers - all operators work directly on memory.
	\begin{itemize}
          \item Clearly impractical to implement in hardware, but this is not a design goal.
	\end{itemize}
\item 2$^{15}$ words, each 68 bits large.
\item 15 operators, each allowing four 16-bit operands.
	\begin{itemize}
          \item Every instruction does not utilize every operand.
          \item The MSB of every operand indicates if the remaining 15 bits refer to an immediate or an address.
	\end{itemize}
\item Program counter stored in a fixed location in memory, allowing the program to modify it (i.e. branching).
\item A special operator \textbf{iterinput} that fetches the next input chunk and places it in a fixed memory location.
\item A fixed section of memory treated as an output buffer.
\end{itemize}

\paragraph{} 
Although the algorithm itself views this language in terms of its machine code, the human-readable disassembly is straightforward, since instructions take the form [ dest ] add op0 [ op2 ]”. The machine code itself is just a series of binary strings. This makes programs trivial to
	\begin{itemize}
	\item Generate: Given a template with gaps, randomly generate bitstrings of length 68 to fill the gaps
	\item Breed: Simply join subsets of the parent chromsomes.
	\item Mutate: AND some random 68-bit number(s) with some random word(s).
	\end{itemize}

@@

The primary performance bottleneck, the fitness testing, is completely parallelized, with a pool of processes that greedily grab functions to simulate and evaluate. Breeding is less of a performance concern, but it is also parallelized due to the ease with which it is possible. There are numerous additional opportunities for algorithmic parallelization that aren't utilized by our implementation.

Special mention must be given to Psyco (http://psyco.sourceforge.net/) and the processing module (http://pypi.python.org/pypi/processing/). Psyco improves individual script execution by utilizing a JIT compiler, while the processing module allows for quick and efficient parallelization of the algorithm itself.

\section{Computational Results}

Unfortunately, the hash functions the algorithm currently generates are very weak, and would at best serve as CRCs or checksums of some sort. In all runs, we detected only minor differences between hash values for lexicographically adjacent inputs. However, we believe this machine learning approach holds promise, and given more time, we feel we can produce more substantial and interesting results. There are two primary reasons. First, the randomized generator function is weak. There are in fact four structured generator functions provided, but only one of them is actually used, since the others never produce anything but constants. The single usable generator is itself a manually-created and very trivial hash function. It simply returns some number of bits from the first block of the input. For small inputs, it is a kind of perfect hash function (http://burtleburtle.net/bob/hash/perfect.html), and obviously produces collisions for longer inputs.

Indeed, on first glance, it would seem like a poor candidate for an initial function. However, we conducted several runs (~50 generations) with the initial population consisting entirely of this function, which is 5 lines long. At the end of each of these runs, we noticed that there were two or three long functions (> 300 lines) whose exact hash outputs varied slightly from their initial forms. The variation wasn't significant, but the fact that phenotypic variation existed at all is a hopeful sign. It stands to reason that if provided additional trivial hash functions that collectively exercised the full power of the assembly language, the generator would allow the algorithm to produce substantially more interesting input.

Secondly, the resources available for the purposes of this paper was limited. We had full-time access to a single dual-core machine that was frequently used for other tasks, and only one time-strapped developer. If the algorithm could be adapted to run on a cluster or in the cloud with the backing of additional human resources, it would allow much more rapid convergence on interesting behavior upon execution.

>>>>

\paragraph{}
The genetic algorithm is inspired by biological natural selection. John Holland of the University of Michigan pioneered its use in the 1970s (Holland, “Genetic algorithms and the optimal allocation of trials.” SIAM J. Comput., 2.88-105); however, it wasn't until the 1980s and 1990s that the computational capacity for practical implementations of the algorithm became readily available. The algorithm as we use it consists of a series of functions - a randomized generator, a fitness evaluator, and a breeder - that operate on chromomsomes, which are strings formed from a genetic description alphabet. The algorithm proceeds as follows:
\begin{enumerate}
	\item Create a random starting population of size S.
	\item Evaluate the fitness of every member of the current generation.
	\item Breed the fittest members to form the next generation.
	\item Add random or old members until we have the next generation is once more of size S.
	\item Go to step 2 if we're not done.
\end{enumerate}

@@

\begin{itemize}
\item No registers - all operators work directly on memory.
	\begin{itemize}
          \item Somewhat impractical to implement in hardware, but this is not a design goal.
	\end{itemize}
\item 2$^{15}$ words of memory, each 68 bits large.
\item 15 operators, each allowing four 16-bit operands.
	\begin{itemize}
          \item Every instruction does not utilize every operand.
          \item Leading bit of every operand indicates if the remaining 15 bits refer to an immediate or an address.
	\end{itemize}
\item Program counter stored in a fixed location in memory, allowing the program to modify it (i.e. branching).
\item A special operator \textbf{iterinput} that fetches the next input chunk and places it in a fixed memory location.
\item A fixed section of memory treated as an output buffer.
\end{itemize}

\paragraph{} 
Although the algorithm itself views this language in terms of its machine code, the human-readable disassembly is straightforward, since instructions take the form \textbf{[dest] [op] [op0] [op2]}. The machine code itself is just a series of binary strings. This makes programs trivial to
	\begin{itemize}
	\item Generate: Given a template with gaps, randomly generate bitstrings of length 68 to fill the gaps
	\item Breed: Simply join subsets of the parent chromsomes.
	\item Mutate: AND some random bitstring(s) of length 68 with some random word(s).
	\end{itemize}

@@

\paragraph{}
The primary performance bottleneck, the fitness testing, is completely parallelized, with a pool of processes that greedily grab functions to simulate and evaluate. Breeding is less of a performance concern, but it is also parallelized due to the ease with which it is possible. There are numerous additional opportunities for algorithmic parallelization that aren't utilized by our implementation.

\paragraph{}
Special mention must be given to Psyco (http://psyco.sourceforge.net/) and the processing module (http://pypi.python.org/pypi/processing/). Psyco improves individual script execution by utilizing a JIT compiler, while the processing module allows for quick and efficient parallelization of the algorithm itself.

\paragraph{}	
The fitness function consists of two main parts. First, it attempts to filter out trivially bad functions by hashing the elements in [-10, 10]. If the function fails this trivial test by colliding at all, it is flagged, and no further tests are conducted on it.
Next, it iterates over many random numbers, some of them very close to each other, to determine collision rates.
The fitness function notes the ratio collisions $\div$ total inputs, and then terminates.
It is important to note that randomness and uniformity tests are not conducted programmatically due to their cost. Over 68 million bytes are required for a single hash function's test. The intended procedure for randomness testing is as follows: 
\begin{enumerate}
\item Manually create the named lockfile in the computational results directory to pause the algorithm.
\item Initiate the randomness tests by hand.
\item Manually append the results of the test to the fitness file of the relevant function. 
\item The results will then be accounted for in the following generation's selection and breeding process.
\end{enumerate}

\section{Computational Results}
Unfortunately, the hash functions the algorithm currently generates are very weak, and would at best serve as CRCs or checksums of some sort. In all runs, we detected only minor differences between hash values for lexicographically adjacent inputs for all generated hash functions. 
\paragraph{}

However, we believe this machine learning approach holds promise, and given more time, we feel we can produce more substantial and interesting results. There are two primary reasons. First, the randomized generator function is weak. There are in fact four structured generator functions provided, but only one of them is actually used, since the others never produce anything but constants. The single usable generator is itself a manually-created and very trivial hash function. It simply returns some number of bits from the first block of the input. For small inputs, it is a kind of perfect hash function (http://burtleburtle.net/bob/hash/perfect.html), and obviously produces collisions for longer inputs.

\paragraph{}
Indeed, on first glance, it would seem like a poor candidate for an initial function. We conducted several runs ($\approx$ 50 generations) with the initial population consisting entirely of this function, which is 5 lines long. At the end of each of these runs, we noticed that there were two or three long functions ($>$300 lines) whose exact hash outputs varied slightly from their initial forms. The variation wasn't significant, but the fact that phenotypic variation existed at all is a hopeful sign. It stands to reason that if provided additional trivial hash functions that collectively exercised the full power of the assembly language, the generator would allow the algorithm to produce substantially more interesting input.

\paragraph{}
Secondly, the resources available for the purposes of this paper were limited. We had full-time access to a single dual-core machine that was frequently used for other tasks, and only one time-strapped developer. With the injection of additional human resources, the algorithm could be adapted to run on a cluster, the cloud, etc. It would definitely allow for more rapid convergence on interesting behavior.
